---
title: "Creating the Dataset"
author: "Jaylen Lee"
date: "10/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
library(lubridate)
```

#Loading in nbastatR
In order to investigate the effects of the bubble environment on shooting, I need to create the appropriate dataset. The nbastatR package by Alex Bresler makes this pretty straightforward. 

```{r}
#If you wanted to install the package you can do so with the following command

#devtools::install_github("abresler/nbastatR") #uncomment to install

#Loading the package
library(nbastatR)
```

Next we need to generate a list of all NBA teams and then exclude the teams that were not invited to the bubble. 

```{r}
#Creating the list of non bubble teams
non_bubble_teams = c("Charlotte Hornets", "Chicago Bulls","New York Knicks",
                     "Detroit Pistons", "Atlanta Hawks", "Cleveland Cavaliers",
                     "Golden State Warriors", "Minnesota Timberwolves")

#Loading the list of teams
current_teams <- nba_teams() %>%
  filter(isNonNBATeam==0) %>%
  mutate(nameTeam = ifelse(nameTeam == "Los Angeles Clippers", "LA Clippers", nameTeam ))

bubble_teams <- current_teams %>%
  filter(!(nameTeam %in% non_bubble_teams))
```


What would also be nice is to have some measure of the defensive strength that is being played against the player. A crude measure is the opponent's defensive rating which basketball reference calculates as the average number of points allowed per 100 possessions. Unfortunately, nbastatR's basketball reference pulling isn't working so I need to try web scraping myself. 

```{r}
#Loading in package for webscraping
library(rvest)

#Opening webpage
br_url <- read_html("https://www.basketball-reference.com/leagues/NBA_2020_ratings.html")

#Grabbing table
team_ratings <- br_url %>% html_nodes("table") %>% html_table(header = T)

#Formatting
team_ratings <-  team_ratings[[1]] #Extracting just the data frame

#Turning first row into a header
names(team_ratings) <- as.character(unlist(team_ratings[1,]))
team_ratings <- team_ratings[-1,]

#Grabbing the columns of interest: Team name and defensive rating
team_ratings <- team_ratings[,c("Team","DRtg")]

#Cleaning
team_ratings$DRtg <- as.numeric(team_ratings$DRtg) #Changing DRtg from character to numeric attribute
names(team_ratings) <- c("nameTeam","defRate") #renaming columns to match previous table

#Final Defensive rating table
team_ratings <- tibble(team_ratings)
team_ratings <- left_join(team_ratings, current_teams[c("nameTeam","slugTeam")], "nameTeam")

#Fixing Clippers Slug
team_ratings[which(team_ratings$nameTeam == "Los Angeles Clippers"), "slugTeam"] = "LAC"
```

#Creating list of shots
Now lets grab all of the shots taken during the 2019-2020 season including the playoffs. We'll also be joining in the defensive rating data and excluding players from teams who did not participate in the bubble.

```{r, message=FALSE, warnings = FALSE}
shotsData <-  teams_shots(teams = bubble_teams$nameTeam,
                          seasons = 2020, 
                          season_types = c("Regular Season", "Playoffs")) %>% #Grabbing data
  
  filter(typeAction == "Jump Shot") %>% #Filtering out non-jumpshots
  
  select(namePlayer, nameTeam, typeEvent, typeShot, dateGame, slugTeamHome, slugTeamAway,
         idGame, distanceShot, zoneBasic, isShotMade,
         locationX, locationY) %>% #Selecting relevant variables
  
  mutate(made_shot = 1*(typeEvent=="Made Shot")) %>% #creating indicator for made shots
  
  left_join(current_teams[c("nameTeam","slugTeam")], "nameTeam" ) %>% #joining in the shortened team names
  
  mutate(home_team = 1*(slugTeam == slugTeamHome)) %>% #creating the hometeam indicator
  
  mutate(opp_slug = ifelse(home_team==1, slugTeamAway, slugTeamHome)) %>% #Creating opponent variable
  
  left_join(team_ratings[c("slugTeam", "defRate")], by = c("opp_slug" = "slugTeam")) #Adding defensive ratings for opponents

#Reformating the date of the game to be as a datetime 
shotsData$dateGame <- ymd(shotsData$dateGame)

#Creating the indicator for whether the game was in the bubble or not
shotsData$bubble <- 1*((shotsData$dateGame > ymd("20200501")) & (shotsData$dateGame < ymd("20201020")))

```

## Saving the dataset
```{r}
#Saving the data
save(shotsData, file = "shot_data.RData")
```



